### 2015-04-14 zoldatoff/tex/text retrieval/lecture08.tex
Лекции по курсу Text Retrieval
Неделя 3. Лекции 6-8

\begin{itemize}
\item Probabilistic models ranking function: 
\begin{equation*}
f(d, q) = p(R=1 \,\big|\, d,q), \; R \in \{0,1\}
\end{equation*}
\item \textbf{Query Likelihood}: if a user likes document d, how likely would the user enter query q (in order to retrieve d)? 
\item Assumption: a user formulates a query based on an <<imaginary relevant document>>:
\begin{equation*}
p(R=1 \,\big|\, d,q) \approx p(q \,\big|\, d,R=1)
\end{equation*}
\item Basic idea based on user clicks (R=1):
\begin{equation*}
f(q, d) = p(R=1 \,\big|\, d,q) = \frac{count(q, d, R=1)}{count(q,d)}
\end{equation*}
\item How to compute $p(q \,|\, d)$? How to compute probability of
text in general? $\to$ \textbf{Language Model}
\end{itemize}

%----------------------------------------
\subsection{Language Model}
The term language model (LM) refers to a probabilistic model of text (i.e., it defines a probability distribution over sequences of words).

@@

%----------------------------------------
\subsection{The Simplest Language Model: Unigram LM}
Unigram Language Model = word distribution
\begin{itemize}
\item Generate text by generating each word \textbf{independently} 
\item Thus, $p(w_1, w_2 \dots w_n) = p(w_1)p(w_2) \dots p(w_n)$
\item Parameters: $\{p(w_i)\}: p(w_1) + \ldots + p(w_N) = 1$ (N is vocabulary size)
\item Text = sample drawn according to this \textbf{word distribution}
\end{itemize}

@@

%----------------------------------------
\subsection{Recommended reading}
\begin{itemize}
\item Chris Manning and Hinrich Sch{\"u}tze, <<Foundations of Statistical Natural Language Processing>>, MIT Press. Cambridge, MA: May 1999.
\item  Rosenfeld, R., <<Two decades of statistical language modeling: where do we go from here?>>, Proceedings of the IEEE , vol.88, no.8, pp.1270,1278, Aug. 2000
\end{itemize}

@@

%----------------------------------------
\subsection{How to Estimate $p(w \big| d)$}
Smoothing of $p(w \big| d)$ is necessary for query likelihood:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{smoothed_lm.png}
\end{figure}

Key Question: what probability should be assigned to an unseen word? Let the probability of an unseen word be proportional to its probability given by a reference LM. One possibility: Reference LM = Collection LM:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{probability_lm.png}
\end{figure}

%----------------------------------------
\subsection{Rewriting the Ranking Function with Smoothing}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{ranking_smoothing.png}
\end{figure}

%----------------------------------------
\subsection{Benefit of Rewriting}

@@

%----------------------------------------
\subsection{Smoothing Methods}
\begin{equation*}
f(q, d) = \sum_{w_i \in d, w_i \in q} c(w_i, q) \log \frac{p_{seen}(w_i \,\big|\, d)}{\alpha_d \, p(w_i \,\big|\, C)} + n \log \alpha_d
\end{equation*}

\subsubsection{Jelinek-Mercer Smoothing}
Jelinek-Mercer: Fixed coefficient linear interpolation
\begin{equation*}
p_{seen}(w \,\big|\, d) = (1-\lambda) \, \dfrac{c(w, d)}{|d|} + \lambda \, p(w \,\big|\, C), \quad \alpha_d = \lambda, \quad \lambda \in [0, 1]
\end{equation*}

Ranking Function:
\begin{equation*}
f_{JM}(q, d) = \sum_{w \in d, w \in q} c(w, q) \log \left( 1 + \frac{1-\lambda}{\lambda} \frac{c(w, d)}{|d| \, p(w \,\big|\, C)} \right)
\end{equation*}

\subsubsection{Dirichlet Prior (Bayesian) Smoothing}
Dirichlet Prior: Adding pseudo counts; adaptive interpolation
\begin{equation*}
p_{seen}(w \,\big|\, d) = \dfrac{c(w, d) + \mu \, p(w \,\big|\, C)}{|d| + \mu} = \dfrac{|d|}{|d| + \mu}\cdot\dfrac{c(w,d)}{|d|} + \dfrac{\mu}{|d| + \mu}\cdot p(w \,\big|\, C)
\end{equation*}

@@

Ranking Function:
\begin{equation*}
f_{DIR}(q, d) = \sum_{w \in d, w \in q} c(w, q) \log \left( 1 + \frac{c(w, d)}{\mu \, p(w \,\big|\, C)} \right) + n\,\log\frac{\mu}{\mu + |d|}
\end{equation*}

%----------------------------------------
\subsection{Summary of Query Likelihood Probabilistic Model}
\begin{itemize}
\item Effective ranking functions obtained using pure probabilistic modeling

\begin{itemize}
\item Assumption 1: $Relevance(q,d) = p(R=1 \,\big|\, q,d) \approx p(q \,\big|\, d,R=1) \approx p(q \,\big|\, d)$
\item Assumption 2: Query words are generated independently
\item Assumption 3: Smoothing with $p(w \,\big|\, C)$
\item Assumption 4: JM or Dirichlet prior smoothing
\end{itemize}

@@

%----------------------------------------
\subsection{Recommended reading}
\begin{itemize}
\item ChengXiang Zhai, \href{http://www.morganclaypool.com/doi/abs/10.2200/S00158ED1V01Y200811HLT001}{<<Statistical Language Models for Information Retrieval>> (Synthesis Lectures Series on Human Language Technologies)}, Morgan \& Claypool Publishers, 2008.
\end{itemize}

>>>>

\begin{itemize}
\item Probabilistic models ranking function: 
\begin{equation*}
f(d, q) = p(R=1 \,\big|\, d,q), \; R \in \{0,1\}
\end{equation*}
\item \textbf{Query Likelihood}: if a user likes document d, how likely would the user enter query q (in order to retrieve d)? 
\item Assumption: a user formulates a query based on an <<imaginary relevant document>>:
\begin{equation*}
p(R=1 \,\big|\, d,q) \approx p(q \,\big|\, d,R=1)
\end{equation*}
\item Basic idea based on user clicks ($R=1$):
\begin{equation*}
f(q, d) = p(R=1 \,\big|\, d,q) = \frac{count(q, d, R=1)}{count(q,d)}
\end{equation*}
\item How to compute $p(q \,|\, d)$? How to compute probability of
text in general? $\to$ \textbf{Language Model}
\end{itemize}

%----------------------------------------
\subsection{Language Model}
The term language model (LM) refers to a probabilistic model of text (i.e., it defines a probability distribution over sequences of words).\\

@@

%----------------------------------------
\subsubsection{The Simplest Language Model: Unigram LM}
Unigram Language Model = word distribution
\begin{itemize}
\item Generate text by generating each word \textbf{independently} 
\item Thus, $p(w_1, w_2 \dots w_n) = p(w_1)p(w_2) \dots p(w_n)$
\item Parameters: $\{p(w_i)\}: p(w_1) + \ldots + p(w_N) = 1$ (N is vocabulary size)
\item Text = sample drawn according to this \textbf{word distribution}
\end{itemize}

@@

%----------------------------------------
\subsubsection{Recommended reading}
\begin{itemize}
\item Chris Manning and Hinrich Sch{\"u}tze, <<Foundations of Statistical Natural Language Processing>>, MIT Press. Cambridge, MA: May 1999.
\item  Rosenfeld, R., <<Two decades of statistical language modeling: where do we go from here?>>, Proceedings of the IEEE , vol.88, no.8, pp.1270,1278, Aug. 2000
\end{itemize}

@@

%----------------------------------------
\subsubsection{How to Estimate $p(w \,\big|\, d)$}
Smoothing of $p(w \big| d)$ is necessary for query likelihood:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{smoothed_lm.png}
\end{figure}

Key question: what probability should be assigned to an unseen word? Let the probability of an unseen word be proportional to its probability given by a reference LM. One possibility: Reference LM = Collection LM:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{probability_lm.png}
\end{figure}

%----------------------------------------
\subsubsection{Rewriting the Ranking Function with Smoothing}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{ranking_smoothing.png}
\end{figure}

@@

%----------------------------------------
\subsubsection{Smoothing Methods}
\begin{equation*}
f(q, d) = \sum_{w_i \in d, w_i \in q} c(w_i, q) \log \frac{p_{seen}(w_i \,\big|\, d)}{\alpha_d \, p(w_i \,\big|\, \mathcal{C})} + n \log \alpha_d
\end{equation*}

\newpage
\paragraph{Jelinek-Mercer Smoothing}\mbox{}\\

Jelinek-Mercer: Fixed coefficient linear interpolation
\begin{equation*}
p_{seen}(w \,\big|\, d) = (1-\lambda) \, \dfrac{c(w, d)}{|d|} + \lambda \, p(w \,\big|\, \mathcal{C}), \quad \alpha_d = \lambda, \quad \lambda \in [0, 1]
\end{equation*}

Ranking Function:
\begin{equation*}
f_{JM}(q, d) = \sum_{w \in d, w \in q} c(w, q) \log \left( 1 + \frac{1-\lambda}{\lambda} \frac{c(w, d)}{|d| \, p(w \,\big|\, \mathcal{C})} \right)
\end{equation*}

\paragraph{Dirichlet Prior (Bayesian) Smoothing}\mbox{}\\

Dirichlet Prior: Adding pseudo counts; adaptive interpolation
\begin{equation*}
p_{seen}(w \,\big|\, d) = \dfrac{c(w, d) + \mu \, p(w \,\big|\, \mathcal{C})}{|d| + \mu} = \dfrac{|d|}{|d| + \mu}\cdot\dfrac{c(w,d)}{|d|} + \dfrac{\mu}{|d| + \mu}\cdot p(w \,\big|\, \mathcal{C})
\end{equation*}

@@

Ranking Function:
\begin{equation*}
f_{DIR}(q, d) = \sum_{w \in d, w \in q} c(w, q) \log \left( 1 + \frac{c(w, d)}{\mu \, p(w \,\big|\, \mathcal{C})} \right) + n\,\log\frac{\mu}{\mu + |d|}
\end{equation*}

%----------------------------------------
\subsubsection{Summary of Query Likelihood Probabilistic Model}
\begin{itemize}
\item Effective ranking functions obtained using pure probabilistic modeling

\begin{itemize}
\item Assumption 1: $Relevance(q,d) = p(R=1 \,\big|\, q,d) \approx p(q \,\big|\, d,R=1) \approx p(q \,\big|\, d)$
\item Assumption 2: Query words are generated independently
\item Assumption 3: Smoothing with $p(w \,\big|\, \mathcal{C})$
\item Assumption 4: JM or Dirichlet prior smoothing
\end{itemize}

@@

%----------------------------------------
\subsubsection{Recommended reading}
\begin{itemize}
\item ChengXiang Zhai, \href{http://www.morganclaypool.com/doi/abs/10.2200/S00158ED1V01Y200811HLT001}{<<Statistical Language Models for Information Retrieval>> (Synthesis Lectures Series on Human Language Technologies)}, Morgan \& Claypool Publishers, 2008.
\end{itemize}

### 2015-04-14 zoldatoff/tex/text retrieval/lecture09.tex
Лекции по курсу Text Retrieval
Неделя 3. Лекции 6-8

\section{Vector Space Retrieval Model: Doc Length Normalization}

\subsection{Pivoted Length Normalization}

\textbf{Pivoted length normalizer}: use average doc length as <<pivot>>\footnote{Pivot - стержень; точка опоры, вращения}. Normalizer = 1 if $\abs{d}$ = average doc length (avdl).

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{pivoted_length_norm.png}
\end{figure}

\subsection{State of the Art VSM Ranking Functions}

Pivoted Length Normalization VSM [Singhal et al 96]:
\begin{equation*}
f(q, d) = \sum_{w \in q \cap d} c(w, q) \frac{\ln[1+\ln(1+c(w, d))]}{1-b+b\frac{|d|}{avdl}} \log \frac{M+1}{df(w)}
\end{equation*}

\href{https://ru.wikipedia.org/wiki/Okapi_BM25}{BM25/Okapi} [Robertson \& Walker 94]:
\begin{equation*}
f(q, d) = \sum_{w \in q \cap d} c(w, q) \frac{(k+1) c(w, d)}{c(w, d) + k\left(1-b+b\frac{|d|}{avdl}\right)} \log \frac{M+1}{df(w)}
\end{equation*}

\subsection{Further Improvement of VSM?}
\begin{itemize}
\item Improved instantiation of dimension?
\begin{itemize}
\item  stemmed words, stop word removal, phrases, latent semantic indexing (word clusters), character n-grams, ...
\item  bag-of-words with phrases is often sufficient in practice
\item  Language-specific and domain-specific tokenization is important to
ensure “normalization of terms”
\end{itemize}

\item  Improved instantiation of similarity function?
\begin{itemize}
\item  cosine of angle between two vectors?
\item  Euclidean?
\item  dot product seems still the best (sufficiently general especially with appropriate term weighting)
\end{itemize}
\end{itemize}

\subsection{Further Improvement of BM25}
\begin{itemize}
\item BM25F [Robertson \& Zaragoza 09]
\begin{itemize}
\item Use BM25 for documents with structures (<<F>>=fields)
\item Key idea: combine the frequency counts of terms in all fields and then apply BM25 (instead of the other way)
\end{itemize}

\item BM25+ [Lv \& Zhai 11]
\begin{itemize}
\item Address the problem of over penalization of long documents
by BM25 by adding a small constant to TF
\item Empirically and analytically shown to be better than BM25
\end{itemize}
\end{itemize}

\subsection{Summary of Vector Space Model}
\begin{itemize}
\item Relevance(q,d) = similarity(q,d)
\item Query and documents are represented as vectors
\item Heuristic\footnote{Heuristic - эвристический} design of ranking function

\item Major term weighting heuristics 
\begin{itemize}
\item TF weighting and transformation 
\item IDF weighting
\item Document length normalization
\end{itemize}

\item BM25 and Pivoted normalization seem to be most effective
\end{itemize}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{pivoted_length_normalization_vsm.png}
\end{figure}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{BM25_Okapi.png}
\end{figure}

\subsection{Recommended reading}
\begin{itemize}
\item A.Singhal, C.Buckley, and M.Mitra. <<Pivoted document length normalization>>. In Proceedings of ACM SIGIR 1996.
\item S. E. Robertson and S. Walker. <<Some simple effective approximations to the 2-Poisson model for probabilistic weighted retrieval>>, Proceedings of ACM SIGIR 1994.
\item S. Robertson and H. Zaragoza. <<The Probabilistic Relevance Framework: BM25 and Beyond>>, Found. Trends Inf. Retr. 3, 4 (April 2009).
\item Y. Lv, C. Zhai, <<Lower-bounding term frequency normalization>>. In Proceedings of ACM CIKM 2011.
\end{itemize}

>>>>

\newpage
\section{Feedback in Text Retrieval}

Feedback = learn from examples

\subsection{Types of Feedback}
\begin{itemize}
\item Relevance Feedback \\ Users make explicit relevance judgments on the initial results. Judgments are reliable, but users don’t want to make extra effort.
\item Pseudo/Blind/Automatic Feedback. \\ Top-k initial results are simply assumed to be relevant. Judgments aren’t reliable, but no user activity is required.
\item Implicit Feedback. \\ User-clicked docs are assumed to be relevant; skipped ones non-relevant. Judgments aren’t completely reliable, but no extra effort from users.
\end{itemize}

%----------------------------------------------
\subsection{Feedback in Vector Space Model}

How can a TR system learn from examples to improve retrieval accuracy? General method - query modification:
\begin{itemize}
\item Adding new (weighted) terms (query expansion)
\item Adjusting weights of old terms
\end{itemize}

%----------------------------------------------
\subsubsection{Rocchio Feedback}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{Rocchio.jpg}
\end{figure}

\begin{equation*}
\overrightarrow{Q_m} = \bigl(\alpha \cdot \overrightarrow{Q_o} \bigr) + \biggl( \frac{\beta}{|D_r|} \cdot \sum_{\overrightarrow{D_j} \in D_r} \overrightarrow{D_j} \biggr)
- \biggl(\frac{\gamma}{|D_{nr}|} \cdot \sum_{\overrightarrow{D_k} \in D_{nr}} \overrightarrow{D_k} \biggr) 
\end{equation*}

\begin{itemize}
\item $\overrightarrow{Q_m}$ - Modified Query Vector
\item $\overrightarrow{Q_o}$ - Original Query Vector
\item $\overrightarrow{D_j}$ - Related Document Vector
\item $\overrightarrow{D_k}$ - Non-Related Document Vector
\item $\alpha$ - Original Query Weight
\item $\beta$ - Related Documents Weight
\item $\gamma$ - Non-Related Documents Weight
\item $D_r$ - Set of Related Documents
\item $D_{nr}$ - Set of Non-Related Documents
\end{itemize}

%----------------------------------------------
\subsubsection{Rocchio in Practice}
\begin{itemize}
\item Negative (non-relevant) examples are not very important
\item Often truncate the vector (i.e., consider only a small number of words that have highest weights in the centroid vector) (efficiency concern)
\item Avoid <<over-fitting>> (keep relatively high weight on the original query weights)
\item Can be used for relevance feedback and pseudo feedback ($\beta$ should be set to a larger value for relevance feedback than for pseudo feedback)
\item Usually robust and effective
\end{itemize}

%----------------------------------------------
\subsection{Feedback in Language Model}

\subsubsection{The KL-divergence measure}
Given two probability mass functions $p(x)$ and $q(x)$, the Kullback-Leibler divergence (or relative entropy) between $p$ and $q$ is defined as:
\begin{equation*}
D(p \parallel q) = \sum_x p(x) \log \frac{p(x)}{q(x)}
\end{equation*}

%--------------
\subsubsection{Kullback-Leibler (KL) Divergence Retrieval Model}

KL-divergence (cross entropy):
\begin{equation*}
f(q, d) = \sum_{\substack{w \in d, \\ p(w \mid \hat{\theta}_Q)>0}} p(w \,\big|\, \hat{\theta}_Q) \log \frac{p_{seen}(w_i \,\big|\, d)}{\alpha_d \, p(w_i \,\big|\, \mathcal{C})} + \log \alpha_d
\end{equation*}
\begin{itemize}
\item $\hat{\theta}_Q$ - estimated query unigram language model
\end{itemize}

If query language model is maximum likelihood $p_{ml}(w \,\big|\, \hat{\theta}_Q) = \dfrac{c(w, q)}{|q|}$, then KL-divergence becomes a query likelihood retrieval formula:
\begin{equation*}
f(q, d) = \sum_{\substack{w_i \in d, \\ w_i \in q}} c(w_i, q) \log \frac{p_{seen}(w_i \,\big|\, d)}{\alpha_d \, p(w_i \,\big|\, \mathcal{C})} + n \log \alpha_d
\end{equation*}

%--------------
\subsubsection{Feedback as Model Interpolation}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{LM_feedback.png}
\end{figure}

\begin{equation*}
p(w \,\big|\, \hat{\theta}_Q) = (1-\alpha) \, p_{ml}(w \,\big|\, \hat{\theta}_Q) + \alpha \, p(w \,\big|\, \theta_F)
\end{equation*}
\begin{itemize}
\item $\alpha$ is a parameter that needs to be set empirically
\item $p(w \,\big|\, \theta_F)$ is a feedback model
\end{itemize}

%--------------
\subsubsection{Generative Mixture Model}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{mixture_model.png}
\end{figure}

\begin{equation*}
\log p(\mathcal{F} \,\big|\, \theta_F) = \sum_{i=1}^n \sum_w c(w, d_i) \log \left( (1-\lambda) p(w \,\big|\, \theta_F) + \lambda p(w \,\big|\, \mathcal{C}) \right)
\end{equation*}
\begin{itemize}
\item $F = \{d_1, \dots , d_n \}$ - the set of feedback documents
\item $\lambda$ - parameter that indicates the amount of <<background noise>> in the feedback documents
\end{itemize}

Maximum likelihood estimate of $\hat{\theta}_F$:
\begin{equation*}
\hat{\theta}_F = \argmax_{\theta_F} \, \log p(\mathcal{F} \,\big|\, \theta_F)
\end{equation*}

%----------------------------------------
\subsection{Recommended reading}
\begin{itemize}
\item ChengXiang Zhai, \href{http://www.morganclaypool.com/doi/abs/10.2200/S00158ED1V01Y200811HLT001}{<<Statistical Language Models for Information Retrieval>> (Synthesis Lectures Series on Human Language Technologies)}, Morgan \& Claypool Publishers, 2008.
\item Victor Lavrenko and W. Bruce Croft. 2001. <<Relevance based language models>>. In Proceedings of ACM SIGIR 2011. 
\end{itemize}
